---
title: "tarea_3"
author: "Pedro Montenegro"
date: "21/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(georob)
library(lattice)
library(dplyr)
```


The package sp provides this data set. According to the help page, it “gives locations and
topsoil heavy metal concentrations, along with a number of soil and landscape variable at
the observation locations, collected in a flood plain of the river Meuse, near the village of
Stein (NL)”.

```{r}
data(meuse, package="sp")
levels(meuse$ffreq) <- paste("ffreq", levels(meuse$ffreq), sep="")
levels(meuse$soil) <- paste("soil", levels(meuse$soil), sep="")
str(meuse)
```


Bivand et al. (2013) use the data to illustrate geostatistical analyses by the package gstat
(Pebesma, 2004). We analyse here the data on zinc in the topsoil (Figure 1).

Figure 1 : Meuse data set: zinc concentration at 155 locations in floodplain of river Meuse near
Stein (NL) shown in Google Earth™.
Figure 1 suggests that zinc concentration depends on distance to the river. We check
graphically whether the two factors ffreq (frequency of flooding) and soil (type) also
influence zinc:


```{r}
palette(trellis.par.get("superpose.symbol")$col)
plot(zinc~dist, meuse, pch=as.integer(ffreq), col=soil)
legend("topright", col=c(rep(1, nlevels(meuse$ffreq)), 1:nlevels(meuse$soil)), pch=c(1:nlevels(meuse$ffreq), rep(1, nlevels(meuse$soil))), bty="n", legend=c(levels(meuse$ffreq), levels(meuse$soil)))
```


Figure 2 : Dependence of zinc concentration on distance to river, frequency of flooding (ffreq)
and soil type.
zinc depends non-linearly on dist and seems in addition to depend on ffreq (larger
concentration at more often flooded sites). Furthermore, the scatter of zinc for given distance increases with decreasing distance (= increasing zinc concentration, heteroscedastic
variation). We use log(zinc) to stabilize the variance:


```{r}
 xyplot(log(zinc)~dist | ffreq, meuse, groups=soil, panel=function(x, y, ...){panel.xyplot(x, y, ...) panel.loess(x, y, ...)}, auto.key=TRUE)

```

Figure 3 : Dependence of zinc on distance to river, frequency of flooding (ffreq) and soil
type.
The relation log(zinc)~dist is still non-linear, hence we transform dist by √ :

```{r}
xyplot(log(zinc)~sqrt(dist) | ffreq, meuse, groups=soil, panel=function(x, y, ...){panel.xyplot(x, y, ...) panel.loess(x, y, ...) panel.lmline(x, y, lty="dashed", ...)}, auto.key=TRUE)
```

Figure 4 : Dependence of zinc concentration on distance to river, frequency of flooding (ffreq)
and soil type.
which approximately linearizes the relation.
The slopes of the regression lines log(zinc)~sqrt(dist) are about the same for all levels
of ffreq. But the intercept of ffreq1 differs from the intercepts of the other levels. Hence,
as an initial drift model we use

```{r}
r.lm <- lm(log(zinc)~sqrt(dist)+ffreq, meuse)
summary(r.lm)

```

The residual diagnostic plots

```{r}
op <- par(mfrow=c(2, 2)); plot(r.lm); par(op)
```


Figure 5 : Residual diagnostic plots for linear drift model log(zinc)~sqrt(dist)+ffreq.
do not show violations of modelling assumptions.
Next, we compute the sample variogram of the residuals for the 4 directions N-S, NE-SW,
E-W, SE-NW by the methods-of-moments estimator:

```{r}
plot(sample.variogram(residuals(r.lm), locations=meuse[, c("x","y")],
lag.dist.def=100, max.lag=2000, xy.angle.def=c(0, 22.5, 67.5, 112.5, 157.5, 180),
estimator="matheron"), type="l",
main="sample variogram of residuals log(zinc)~sqrt(dist)+ffreq")
```


Figure 6 : Direction-dependent sample variogram of regression residuals of
log(zinc)~sqrt(dist)+ffreq.
The residuals appear to be spatially dependent. For the short lags there is no clear
dependence on direction, hence, we assume that auto-correlation is isotropic.
To complete the exploratory modelling exercise, we compute the direction-indepdendent
sample variogram and fit a spherical variogram model by weighted non-linear least squares
(using Cressie’s weights)

```{r}
plot(r.sv <- sample.variogram(residuals(r.lm), locations=meuse[, c("x","y")],
lag.dist.def=100, max.lag=2000,
estimator="matheron"), type="l",
main="sample variogram of residuals log(zinc)~sqrt(dist)+ffreq")

lines(r.sv.spher <- fit.variogram.model(r.sv, variogram.mode="RMspheric",
param=c(variance=0.1, nugget=0.05, scale=1000)))
```

Figure 7 : Sample variogram of regression residuals of log(zinc)~sqrt(dist)+ffreq along with fitted spherical variogram function. and output the fitted variogram parameters

```{r}
summary(r.sv.spher)
```


3.2 Fitting a spatial linear model by Gaussian (RE)ML
We fit the model that we developed in the exploratory analysis now by Gaussian REML:

```{r}
 r.georob.m0.spher.reml <- georob(log(zinc)~sqrt(dist)+ffreq, meuse, locations=~x+y,
+ variogram.model="RMspheric", param=c(variance=0.1, nugget=0.05, scale=1000),
+ tuning.psi=1000)

summary(r.georob.m0.spher.reml)

```

The diagnostics at the begin of the summary output suggest that maximization of the
restricted log-likelihood by nlminb() was successful. Nevertheless, before we interprete
the output, we compute the profile log-likelihood for the range to see whether the maximization has found the global maximum:

```{r}
 r.prfl.m0.spher.reml.scale <- profilelogLik(r.georob.m0.spher.reml,
values=data.frame(scale=seq(500, 5000, by=50)))
plot(loglik~scale, r.prfl.m0.spher.reml.scale, type="l")
abline(v=coef(r.georob.m0.spher.reml, "variogram")["scale"], lty="dashed")
abline(h=r.georob.m0.spher.reml$loglik - 0.5*qchisq(0.95, 1), lty="dotted")
```


Figure 8 : Restricted profile log-likelihood for range parameter (scale) of spherical variogram
(vertical line: estimate of scale returned by georob(); intersection of horizontal line with profile
defines a 95% confidence region for scale based on likelihood ratio test).
Although the restricted log-likelihood is multimodal — which is often observed for variogram models with compact support — we were lucky to find the global maximum because
the initial values of the variogram parameters were close to the REML estimates. Estimates of scale (range of variogram) and variance (partial sill) are correlated, nugget
and scale less so:


```{r}
op <- par(mfrow=c(1,2), cex=0.66)
plot(variance~scale, r.prfl.m0.spher.reml.scale, ylim=c(0, max(variance)), type="l")
plot(nugget~scale, r.prfl.m0.spher.reml.scale, ylim=c(0, max(nugget)), type="l")
par(op)

```

Figure 9 : Estimates of variance (partial sill) and nugget as a function of the estimate of the
range (scale) of the variogram.
We now study the summary output in detail: The estimated variogram parameters are
reported along with 95% confidence intervals that are computed based on the asymptotic
normal distribution of (RE)ML estimates from the observed Fisher information.
The dependence of log(zinc) on sqrt(dist) is highly significant, as is the dependence
on ffreq:

```{r}
waldtest(r.georob.m0.spher.reml, .~.-ffreq)
```

We can test equality of all pairs of intercepts by functions of the package multcomp


```{r}
library(multcomp)
summary(glht(r.georob.m0.spher.reml,
linfct = mcp(ffreq = c("ffreq1 - ffreq2 = 0", "ffreq1 - ffreq3 = 0",
"ffreq2 - ffreq3 = 0"))))
```


As suspected only the intercept of ffreq1 differs from the others. Adding the interaction
sqrt(dist):ffreq does not improve the model:

```{r}
waldtest(r.georob.m0.spher.reml, .~.+sqrt(dist):ffreq)
```


```{r}
waldtest(r.georob.m0.spher.reml, .~.+soil)
```

Drift models may also be build by step-wise covariate selection based on AIC, either
keeping the variogram parameters fixed (default) or re-estimating them afresh for each evaluated model


```{r}
step(r.georob.m0.spher.reml, scope=log(zinc)~ffreq*sqrt(dist)+soil)
```

```{r}
step(r.georob.m0.spher.reml, scope=log(zinc)~ffreq*sqrt(dist)+soil,
+ fixed.add1.drop1=FALSE)
```

which selects the same model. Note that step-wise covariate selection by step.georob(),
add1.georob() and drop1.georob() requires the non-restricted log-likelihood. Before
evaluating candidate models, the initial model is therefore re-fitted by ML, which can be
done by

```{r}
r.georob.m0.spher.ml <- update(r.georob.m0.spher.reml,
control=control.georob(ml.method="ML"))

extractAIC(r.georob.m0.spher.reml, REML=TRUE)
```

```{r}
extractAIC(r.georob.m0.spher.ml)

r.georob.m0.spher.ml
```


Models can be also compared by cross-validation


```{r}
r.cv.m0.spher.reml <- cv(r.georob.m0.spher.reml, seed=3245, lgn=TRUE)
r.georob.m1.spher.reml <- update(r.georob.m0.spher.reml, .~.-ffreq)
r.cv.m1.spher.reml <- cv(r.georob.m1.spher.reml, seed=3245, lgn=TRUE)
summary(r.cv.m0.spher.reml)
```

```{r}
summary(r.cv.m1.spher.reml)
```

Note that the argument lng=TRUE has the effect that the cross-validation predictions of a
log-transformed response are transformed back to the original scale of the measurements.

```{r}
op <- par(mfrow=c(3,2))
plot(r.cv.m1.spher.reml, "sc")
plot(r.cv.m0.spher.reml, "sc", add=TRUE, col=2)
abline(0, 1, lty="dotted")
legend("topleft", pch=1, col=1:2, bty="n",
legend=c("log(zinc)~sqrt(dist)", "log(zinc)~sqrt(dist)+ffreq"))
plot(r.cv.m1.spher.reml, "lgn.sc"); plot(r.cv.m0.spher.reml, "lgn.sc", add=TRUE, col=2)
abline(0, 1, lty="dotted")
plot(r.cv.m1.spher.reml, "hist.pit")
plot(r.cv.m0.spher.reml, "hist.pit", col=2)
plot(r.cv.m1.spher.reml, "ecdf.pit")
plot(r.cv.m0.spher.reml, "ecdf.pit", add=TRUE, col=2)
abline(0, 1, lty="dotted")
plot(r.cv.m1.spher.reml, "bs")
plot(r.cv.m0.spher.reml, add=TRUE, "bs", col=2)
par(op)
```

Figure 10 : Diagnostic plots of cross-validation predictions of REML fits of models
log(zinc)~sqrt(dist) (blue) and log(zinc)~sqrt(dist)+ffreq (magenta).
The simpler model gives less precise predictions (larger rmse, Brier score and therefore
also larger continuous ranked probability score crps), but it models prediction uncertainty
better (PIT closer to uniform distribution, see section 7.3 and Gneiting et al., 2007).
We finish modelling by plotting residual diagnostics of the model
r.georob.m0.spher.reml and comparing the estimated variogram with the ML
estimate and the model fitted previously to the sample variogram of ordinary least
squares (OLS) residuals:

```{r}
op <- par(mfrow=c(2,2), cex=0.66)
plot(r.georob.m0.spher.reml, "ta"); abline(h=0, lty="dotted")
plot(r.georob.m0.spher.reml, "qq.res"); abline(0, 1, lty="dotted")
plot(r.georob.m0.spher.reml, "qq.ranef"); abline(0, 1, lty="dotted")
plot(r.georob.m0.spher.reml, lag.dist.def=100, max.lag=2000)
lines(r.georob.m0.spher.ml, col=2); lines(r.sv.spher, col=3)
par(op)
```

Figure 11 : Residual diagnostic plots of model log(zinc)~sqrt(dist)+ffreq and spherical
variogram estimated by REML (blue), ML (magenta) and fit to sample variogram (darkgreen).
As expected REML estimates larger range and partial sill parameters. The diagnostics
plots do not reveal serious violations of modelling assumptions.

