---
title: "tarea_3"
author: "Pedro Montenegro"
date: "21/10/2019"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(georob)
library(multcomp)
library(lattice)
library(dplyr)
library(broom)
library(ggplot2)
library(ggfortify)
```

## Análisis Geoestadístico análisis del modelo Gausiano para concentraciones de zinc

Los datos meuse contienen información sobre la ubicación y concentraciones de metales pesados en la capa superior del suelo, junto con una serie de variables de suelo y paisaje en los lugares de observación, recogidas en una llanura del río Meuse, cerca del pueblo de Stein.

```{r}
data(meuse, package="sp")
levels(meuse$ffreq) <- paste("ffreq", levels(meuse$ffreq), sep="")
levels(meuse$soil) <- paste("soil", levels(meuse$soil), sep="")
#glimpse(meuse)
```

# Análisis exploratorio

Se presenta por medio de un mapa las distribución de concentracion de zinc en Meuse por medio de 155 ubicaciones, identificadas en una escala que representa mayor concentración del zinc a colores más rojos y menor concentración a colores más claros.

Figura 1: Distribución de zinc en Meuse. Tomado tutorial y manual análisis geoestadistica con r paquete georob. 

En el mapa se hay indicios de la dependencia entre la concentración de zinc y la distancia al río. Por tanto, se valida gráficamente mediante los factores de frecuencias de inundación y el tipo de suelo, mismo que se muestra en el siguiente gráfico.

```{r}
palette(trellis.par.get("superpose.symbol")$col)
plot(zinc~dist, meuse, pch=as.integer(ffreq), col=soil)
legend("topright", col=c(rep(1, nlevels(meuse$ffreq)), 1:nlevels(meuse$soil)), pch=c(1:nlevels(meuse$ffreq), rep(1, nlevels(meuse$soil))), bty="n", legend=c(levels(meuse$ffreq), levels(meuse$soil)))
```


Figure 2 : Dependence of zinc concentration on distance to river, frequency of flooding (ffreq) and soil type.


el zinc depende no linealmente de dist y parece además depender de ffreq (mayor concentración en sitios inundados con mayor frecuencia). Además, la dispersión de zinc para una distancia dada aumenta con la distancia decreciente (= aumento de la concentración de zinc, variación heteroscedastica). Usamos log (zinc) para estabilizar la varianza:

```{r}
xyplot(log(zinc)~dist | ffreq, meuse, groups=soil, panel=function(x, y, ...){
panel.xyplot(x, y, ...)
panel.loess(x, y, ...)
   }, auto.key=TRUE)

```

Figure 3 : Dependence of zinc on distance to river, frequency of flooding (ffreq) and soil
type.
The relation log(zinc)~dist is still non-linear, hence we transform dist by √ :

```{r}
xyplot(log(zinc)~sqrt(dist) | ffreq, meuse, groups=soil, panel=function(x, y, ...){
  panel.xyplot(x, y, ...) 
  panel.loess(x, y, ...) 
  panel.lmline(x, y, lty="dashed", ...)
  }, auto.key=TRUE)
```

Figure 4 : Dependence of zinc concentration on distance to river, frequency of flooding (ffreq)
and soil type.
which approximately linearizes the relation.
The slopes of the regression lines log(zinc)~sqrt(dist) are about the same for all levels
of ffreq. But the intercept of ffreq1 differs from the intercepts of the other levels. Hence,
as an initial drift model we use

```{r}
r.lm <- lm(log(zinc)~sqrt(dist)+ffreq, meuse)
tidy(summary(r.lm))
```

Diagnostico de los gráficos de residuos

```{r}
#op <- par(mfrow=c(2, 2)); plot(r.lm); par(op)
autoplot(r.lm)
```


Los gráficos de residuos no muestran violación de los supuestos, en los residuos en comparación con los ajustados se comprueba que los residuos se distribuyen aleatoriamente con varianza constante, de la misma forma los residuos estandarizados presentan una distribución normal observada en el gráfico de normalidad.

En el caso de los valores influenciales evaluados por las distancias de cooks se concentra la mayor parte entre -2 y 2, con pocos casos que son atípicos e influenciales.


Next, we compute the sample variogram of the residuals for the 4 directions N-S, NE-SW,
E-W, SE-NW by the methods-of-moments estimator:

```{r}
plot(sample.variogram(residuals(r.lm), locations=meuse[, c("x","y")],
lag.dist.def=100, max.lag=2000, xy.angle.def=c(0, 22.5, 67.5, 112.5, 157.5, 180),
estimator="matheron"), type="l",
main="Muestra de los residuos del variograma log(zinc)~sqrt(dist)+ffreq")
```


Figure 6 : Direction-dependent sample variogram of regression residuals of log(zinc)~sqrt(dist)+ffreq.

The residuals appear to be spatially dependent. For the short lags there is no clear dependence on direction, hence, we assume that auto-correlation is isotropic.

To complete the exploratory modelling exercise, we compute the direction-indepdendent sample variogram and fit a spherical variogram model by weighted non-linear least squares (using Cressie’s weights)

```{r}
plot(r.sv <- sample.variogram(residuals(r.lm), locations=meuse[, c("x","y")],
lag.dist.def=100, max.lag=2000,
estimator="matheron"), type="l",
main="Muestra de los residuos del variograma log(zinc)~sqrt(dist)+ffreq")

lines(r.sv.spher <- fit.variogram.model(r.sv, variogram.mode="RMspheric",
param=c(variance=0.1, nugget=0.05, scale=1000)))
```

Figure 7 : Sample variogram of regression residuals of log(zinc)~sqrt(dist)+ffreq along with fitted spherical variogram function. and output the fitted variogram parameters

```{r}
a <-(summary(r.sv.spher))

tidy((a$param.aniso)[[1]])

rsq<-(a$sse);rsq
```


# 3.2 Fitting a spatial linear model by Gaussian (RE)ML

We fit the model that we developed in the exploratory analysis now by Gaussian REML:

```{r}
r.georob.m0.spher.reml <- georob(log(zinc)~sqrt(dist)+ffreq, meuse, locations=~x+y,
variogram.model="RMspheric", param=c(variance=0.1, nugget=0.05, scale=1000),
tuning.psi=1000)

summary(r.georob.m0.spher.reml)

```

The diagnostics at the begin of the summary output suggest that maximization of the restricted log-likelihood by nlminb() was successful. Nevertheless, before we interprete the output, we compute the profile log-likelihood for the range to see whether the maximization has found the global maximum:

```{r}
r.prfl.m0.spher.reml.scale <- profilelogLik(r.georob.m0.spher.reml,
values=data.frame(scale=seq(500, 5000, by=50)))
plot(loglik~scale, r.prfl.m0.spher.reml.scale, type="l")
abline(v=coef(r.georob.m0.spher.reml, "variogram")["scale"], lty="dashed")
abline(h=r.georob.m0.spher.reml$loglik - 0.5*qchisq(0.95, 1), lty="dotted")
```


Figure 8 : Restricted profile log-likelihood for range parameter (scale) of spherical variogram (vertical line: estimate of scale returned by georob(); intersection of horizontal line with profile defines a 95% confidence region for scale based on likelihood ratio test).

Although the restricted log-likelihood is multimodal — which is often observed for variogram models with compact support — we were lucky to find the global maximum because the initial values of the variogram parameters were close to the REML estimates. Estimates of scale (range of variogram) and variance (partial sill) are correlated, nugget and scale less so:

```{r}
op <- par(mfrow=c(1,2), cex=0.66)
plot(variance~scale, r.prfl.m0.spher.reml.scale, ylim=c(0, max(variance)), type="l")
plot(nugget~scale, r.prfl.m0.spher.reml.scale, ylim=c(0, max(nugget)), type="l")
par(op)

```

Figure 9 : Estimates of variance (partial sill) and nugget as a function of the estimate of the range (scale) of the variogram.

We now study the summary output in detail: The estimated variogram parameters are reported along with 95% confidence intervals that are computed based on the asymptotic normal distribution of (RE)ML estimates from the observed Fisher information. The dependence of log(zinc) on sqrt(dist) is highly significant, as is the dependence on ffreq:

```{r}
waldtest(r.georob.m0.spher.reml, .~.-ffreq)
```

We can test equality of all pairs of intercepts by functions of the package multcomp

```{r}
summary(glht(r.georob.m0.spher.reml,
linfct = mcp(ffreq = c("ffreq1 - ffreq2 = 0", "ffreq1 - ffreq3 = 0",
"ffreq2 - ffreq3 = 0"))))
```

As suspected only the intercept of ffreq1 differs from the others. Adding the interaction sqrt(dist):ffreq does not improve the model:

```{r}
waldtest(r.georob.m0.spher.reml, .~.+sqrt(dist):ffreq)
```


```{r}
waldtest(r.georob.m0.spher.reml, .~.+soil)
```

Drift models may also be build by step-wise covariate selection based on AIC, either keeping the variogram parameters fixed (default) or re-estimating them afresh for each evaluated model

```{r}
step(r.georob.m0.spher.reml, scope=log(zinc)~ffreq*sqrt(dist)+soil)
```

```{r}
step(r.georob.m0.spher.reml, scope=log(zinc)~ffreq*sqrt(dist)+soil,
fixed.add1.drop1=FALSE)
```

which selects the same model. Note that step-wise covariate selection by step.georob(),
add1.georob() and drop1.georob() requires the non-restricted log-likelihood. Before
evaluating candidate models, the initial model is therefore re-fitted by ML, which can be done by

```{r}
r.georob.m0.spher.ml <- update(r.georob.m0.spher.reml,
control=control.georob(ml.method="ML"))

extractAIC(r.georob.m0.spher.reml, REML=TRUE)
```

```{r}
extractAIC(r.georob.m0.spher.ml)

r.georob.m0.spher.ml
```


Models can be also compared by cross-validation


```{r}
r.cv.m0.spher.reml <- cv(r.georob.m0.spher.reml, seed=3245, lgn=TRUE)
r.georob.m1.spher.reml <- update(r.georob.m0.spher.reml, .~.-ffreq)
r.cv.m1.spher.reml <- cv(r.georob.m1.spher.reml, seed=3245, lgn=TRUE)
summary(r.cv.m0.spher.reml)
```

```{r}
summary(r.cv.m1.spher.reml)
```

Note that the argument lng=TRUE has the effect that the cross-validation predictions of a
log-transformed response are transformed back to the original scale of the measurements.

```{r}
op <- par(mfrow=c(3,2))
plot(r.cv.m1.spher.reml, "sc")
plot(r.cv.m0.spher.reml, "sc", add=TRUE, col=2)
abline(0, 1, lty="dotted")
legend("topleft", pch=1, col=1:2, bty="n",
legend=c("log(zinc)~sqrt(dist)", "log(zinc)~sqrt(dist)+ffreq"))
plot(r.cv.m1.spher.reml, "lgn.sc"); plot(r.cv.m0.spher.reml, "lgn.sc", add=TRUE, col=2)
abline(0, 1, lty="dotted")
plot(r.cv.m1.spher.reml, "hist.pit")
plot(r.cv.m0.spher.reml, "hist.pit", col=2)
plot(r.cv.m1.spher.reml, "ecdf.pit")
plot(r.cv.m0.spher.reml, "ecdf.pit", add=TRUE, col=2)
abline(0, 1, lty="dotted")
plot(r.cv.m1.spher.reml, "bs")
plot(r.cv.m0.spher.reml, add=TRUE, "bs", col=2)
par(op)
```

Figure 10 : Diagnostic plots of cross-validation predictions of REML fits of models
log(zinc)~sqrt(dist) (blue) and log(zinc)~sqrt(dist)+ffreq (magenta).
The simpler model gives less precise predictions (larger rmse, Brier score and therefore
also larger continuous ranked probability score crps), but it models prediction uncertainty
better (PIT closer to uniform distribution, see section 7.3 and Gneiting et al., 2007).
We finish modelling by plotting residual diagnostics of the model
r.georob.m0.spher.reml and comparing the estimated variogram with the ML
estimate and the model fitted previously to the sample variogram of ordinary least
squares (OLS) residuals:

```{r}
op <- par(mfrow=c(2,2), cex=0.66)
plot(r.georob.m0.spher.reml, "ta"); abline(h=0, lty="dotted")
plot(r.georob.m0.spher.reml, "qq.res"); abline(0, 1, lty="dotted")
plot(r.georob.m0.spher.reml, "qq.ranef"); abline(0, 1, lty="dotted")
plot(r.georob.m0.spher.reml, lag.dist.def=100, max.lag=2000)
lines(r.georob.m0.spher.ml, col=2); lines(r.sv.spher, col=3)
par(op)
```

Figure 11 : Residual diagnostic plots of model log(zinc)~sqrt(dist)+ffreq and spherical
variogram estimated by REML (blue), ML (magenta) and fit to sample variogram (darkgreen).
As expected REML estimates larger range and partial sill parameters. The diagnostics
plots do not reveal serious violations of modelling assumptions.

